A lottery model of density-dependent selection

### 2017-10-28 manuscript.tex
latest

MacArthur famously distinguished between ``$r$-selection'' for faster growth at low population density, and ``$K$-selection'' for the ability to keep growing at high densities \citep{macarthur_1962,macarthur_1967}. While the $r$/$K$ scheme is primarily known for its seminal role in the study of life-history evolution \citep{pianka_1970,boyce_1984,charlesworth_1994,reznick_2002}, it
also initiated a sizeable literature on density-dependent selection. This literature is remarkable for containing some of earliest models of the interplay between ecology and evolution \citep{roughgarden_1979,travis_2013}.

Verbal interpretations of the $r$/$K$ scheme, and the widespread use of the logistic model when discussing it, have tended to obscure the compelling generality of MacArthur's argument. We therefore  briefly review that argument now. Consider two types with densities $n_1$ and $n_2$ subject to density-dependent population regulation described by
\begin{equation}
\frac{d n_1}{d t}=f_1(n_1,n_2)\qquad\frac{d n_2}{d t}=f_2(n_1,n_2), \label{eq:macgeneral}
\end{equation}
where $f_1$ and $f_2$ are strictly decreasing functions of $n_1$ and $n_2$. By plotting the nullclines $f_1(n_1,n_2)=0$ and $f_2(n_1,n_2)=0$ in $(n_1,n_2)$ space, it becomes clear that for type $2$ to persist, its nullcline must enclose a region of $(n_1,n_2)$ space that type $1$'s nullcline does not. Since $f_1$ and $f_2$ are strictly decreasing, this corresponds to growth of type $2$ in a region of greater $n_1$ and/or $n_2$ than can be tolerated by type $1$. The term ``$K$-selection'' comes from four ``carrying capacities'' defined as the points where the nullclines intersect the axes (i.e. $f_1(K_{11},0)=0$, $f_1(0,K_{12})=0$, $f_2(0,K_{22})=0$, $f_2(K_{21},0)=0$). These intersection points summarize the nullcline structure provided that the nullclines are sufficiently linear, leading to the conclusion that in ``crowded populations ... fitness is $K$'' \citep[pp. 149]{macarthur_1967}

Confusingly, only $K_{11}$ and $K_{22}$ are saturation densities; $K_{12}$ and $K_{21}$ measure the effects of competition between types. For instance, under Lotka-Volterra competition we have $f_1(n_1,n_2)=r_1(1-\alpha_{11}n_1-\alpha_{12}n_2)n_1$ and $K_{12}=1/\alpha_{12}$, where $\alpha_{12}$ is the per-capita competitive effect of the second type on the first. Thus, we can distinguish between selection for greater saturation density and selection for greater competitive ability (``$\alpha$-selection'') \citep{gill_1974,case_1974,joshi_2001}. Nevertheless, it seems hard not to agree that ``fitness is $K$'' when interpreted in the broad sense that selection at high density either favors the ability to keep growing at ever greater densities, or the ability to stop your competitors' growth at lower densities.

>>>>

MacArthur famously distinguished between ``$r$-selection'' for faster growth at low population density, and ``$K$-selection'' for the ability to keep growing at high densities \citep{macarthur_1962,macarthur_1967}. While the $r$/$K$ scheme is primarily known in the life-history context \citep{pianka_1970,boyce_1984,charlesworth_1994,reznick_2002}, it
also initiated a sizeable literature on density-dependent selection. This literature is remarkable for containing some of earliest models of the interplay between ecology and evolution \citep{roughgarden_1979,travis_2013}.

Verbal interpretations of the $r$/$K$ scheme, and the widespread use of the logistic model when discussing it, have tended to obscure the compelling generality of MacArthur's argument. We therefore  briefly review that argument now. Consider two types with densities $n_1$ and $n_2$ subject to density-dependent population regulation described by
\begin{equation}
\frac{d n_1}{d t}=f_1(n_1,n_2)\qquad\frac{d n_2}{d t}=f_2(n_1,n_2), \label{eq:macgeneral}
\end{equation}
where $f_1$ and $f_2$ are strictly decreasing functions of $n_1$ and $n_2$. By plotting the nullclines $f_1(n_1,n_2)=0$ and $f_2(n_1,n_2)=0$ in $(n_1,n_2)$ space, it becomes clear that for type $2$ to persist, its nullcline must enclose a region of $(n_1,n_2)$ space that type $1$'s nullcline does not. Since $f_1$ and $f_2$ are strictly decreasing, this corresponds to growth of type $2$ in a region of greater $n_1$ and/or $n_2$ than can be tolerated by type $1$. The term ``$K$-selection'' comes from four ``carrying capacities'' defined as the points where the nullclines intersect the axes (i.e. $f_1(K_{11},0)=0$, $f_1(0,K_{12})=0$, $f_2(0,K_{22})=0$, $f_2(K_{21},0)=0$). These intersection points summarize the nullcline structure provided that the nullclines are sufficiently linear, and led MacArthur to conclude that in ``crowded populations ... fitness is $K$'' \citep[pp. 149]{macarthur_1967}

Note that only $K_{11}$ and $K_{22}$ conform with the widespread logistic interpretation of ``$K$'' as representing saturation density; $K_{12}$ and $K_{21}$ measure the effects of competition between types. For instance, under Lotka-Volterra competition we have $f_1(n_1,n_2)=r_1(1-\alpha_{11}n_1-\alpha_{12}n_2)n_1$ and $K_{12}=1/\alpha_{12}$, where $\alpha_{12}$ is the per-capita competitive effect of the second type on the first. To avoid this confusion we can distinguish selection for greater saturation density and selection for greater competitive ability (``$\alpha$-selection'') \citep{gill_1974,case_1974,joshi_2001}. Nevertheless, it seems hard not to agree that ``fitness is $K$'' when interpreted in the broad sense that selection at high density either favors the ability to keep growing at ever greater densities, or the ability to stop your competitors' growth at lower densities.
